---
title: "Investigating IMDB Runtime"
author: "Luke Hannan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::github_document:
    html_preview: false
    toc: true
    toc_depth: 2
# html_document:
#   df_print: paged
#   highlight: pygments
#   code_folding: show
#   toc: yes
#   toc_float: TRUE
knit: (
  function(inputFile, encoding) { 
  
    rmarkdown::render( 
      input       = inputFile, 
      encoding    = encoding, 
      output_file = "README") })
---

# Setup
```{r setup, message = FALSE, warning = FALSE}
library("knitr")
knitr::opts_knit$set(dev = "svg", dev.args = list(type = "cairo", bg = "transparent"))
knitr::opts_chunk$set(dev = "svg", dev.args = list(bg = "transparent"), dpi = 300, warning = FALSE)

library("data.table")
library("parallel")
# for string manipulation
library("stringr")
# for plotting
library("ggplot2")
# for making better looking default ggplot themes
library("ggthemes")
# for more ggplot2 colour schemes
library("colorspace")
# for arranging facet plots
library("cowplot")
# for extra publication related things
library("ggpubr")
# lollipop plots; Author: https://github.com/hrbrmstr
# I just can't install the proj package so I cant install ggalt which has the code for lollipop plots
# so I have just copied that specific geom
source("utils/geom_lollipop.r")

```

# Background

I enjoy watching movies of all sorts, but the past couple of years I have found myself saying **"I enjoyed the movie, but I would have enjoyed it more if they cut a bit of stuff out"** quite a lot. I felt like the average runtime of movies had increased over the years. So I did what any reasonable person would do and looked at the data.

## Data Sources
All data was retrieved from IMDB website on 30-04-2022.

https://datasets.imdbws.com/

Data Documentation: http://www.imdb.com/interfaces/

**title.basics.tsv.gz - Contains the following information for titles:**\
    tconst (string) - alphanumeric unique identifier of the title\
    titleType (string) – the type/format of the title (e.g. movie, short, tvseries, tvepisode, video, etc)\
    primaryTitle (string) – the more popular title / the title used by the filmmakers on promotional materials at the point of release\
    originalTitle (string) - original title, in the original language\
    isAdult (boolean) - 0: non-adult title; 1: adult title\
    startYear (YYYY) – represents the release year of a title. In the case of TV Series, it is the series start year\
    endYear (YYYY) – TV Series end year. ‘\N’ for all other title types\
    runtimeMinutes – primary runtime of the title, in minutes\
    genres (string array) – includes up to three genres associated with the title\

# Data Processing
```{r data_ingest}

# we check if the datafile exists if it does we carry on if it doesn't we download it
# link for the datafiles: https://datasets.imdbws.com/title.basics.tsv.gz
# I don't include it automatically because I don't want to track changes on the dataset or store it in the repo
# If you want to rerun this document first download the data

# The title.basics.tsv.gz file contains the runtime, title, and year of release' exactly what we are looking for
movies = fread("data/title.basics.tsv.gz", sep = r"{\t}", nThread = detectCores(), na.strings = c(",,", r"{\N}"))

# convert runtime to numeric
movies[, runtimeMinutes := as.numeric(runtimeMinutes)]
movies[, startYear := as.numeric(startYear)]

# Then only include movies that have runtimes
movies = subset(movies, titleType == "movie" & !is.na(runtimeMinutes) & !is.na(startYear))
# 367 883 movies that have runtimes
dim(movies)

# First let us see the range of years we have to work with
summary(movies$startYear, na.rm = TRUE)
summary(movies$runtimeMinutes, na.rm = TRUE)
hist(movies$runtimeMinutes)

# some strange runtime and year values we will have to investigate and remove

# lets see these movies with strange runtimes
movies[runtimeMinutes >= 1000, ]

# I'm not interested in any movies that have not been released so any movie that has a startYear after 2021
# for now lets keep these runtimeMinutes outliers in, they wont change much and can always remove them later
movies = subset(movies, startYear <= 2021 & startYear >= 1900)
```
# Overall Movies
Below is a plot of the average run time for all movies in the database from the year 1900 to the year 2020.

```{r overall, out.width = '100%', class.source = 'fold-hide'}
ggplot(aes(x = startYear, y = mean_runtime, colour = n), data = movies[, .(mean_runtime = mean(runtimeMinutes), n = .N), by = startYear]) +
  geom_line(size = 2, lineend = "round", linejoin = "mitre") +
  geom_hline(alpha = 0.8, yintercept = 89.75, linetype = 2, size = 1.15, color = "darkgrey") +
  scale_y_continuous(limits = c(0, 120), breaks = c(0, 30, 60, 90, 120), expand = c(0, 5)) +
  scale_x_continuous(expand = c(0, 1)) +
  scale_colour_gradient(
    low = "#E7B800", high = "#FC4E07", breaks = c(0, 4000, 8000, 12000),
    guide = guide_colorbar(
      title.position = "top", title.hjust = 0.05, nbin = 500, barheight = 0.45, barwidth = 8,
      raster = TRUE
    )
  ) +
  labs(
    x = "Year", y = "Mean Runtime (in Minutes)", colour = "Number of Movies", title = "Mean runtime for all movies from
    1900 to 2020", subtitle = "Dotted line is overall mean runtime (89.75 min). Data from IMDB"
  ) +
  theme_minimal() +
  theme(
    legend.position = c(0.75, 0.25),
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
    text = element_text(size = 12),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold")
  )
```

Lets zoom in real quick and only look at movies that have been released since I was born. Interestingly enough the line seems completely flat, lets throw in a pearsons correlation coefficient to see how flat it is. 
```{r overall_slice, out.width = '100%', class.source = 'fold-hide'}
ggplot(aes(x = startYear, y = mean_runtime, colour = n), data = movies[startYear >= 1994, .(mean_runtime = mean(runtimeMinutes), n = .N), by = startYear]) +
  geom_line(size = 2, lineend = "round", linejoin = "mitre") +
  geom_hline(alpha = 0.8, yintercept = 89.75, linetype = 2, size = 1.15, color = "darkgrey") +
  scale_y_continuous(limits = c(0, 120), breaks = c(0, 30, 60, 90, 120), expand = c(0, 5)) +
  scale_x_continuous(expand = c(0, 1)) +
  scale_colour_gradient(
    low = "#E7B800", high = "#FC4E07", breaks = c(0, 4000, 8000, 12000), guide = guide_colorbar(
      title.position = "top",
      title.hjust = 0.05, nbin = 500, barheight = 0.45, barwidth = 8, raster = TRUE
    )
  ) +
  labs(
    x = "Year", y = "Mean Runtime (in Minutes)", colour = "Number of Movies", title = "Mean runtime for all movies from
    1994 to 2020", subtitle = "Dotted line is overall mean runtime (89.75 min). Data from IMDB"
  ) +
  theme_minimal() +
  theme(
    legend.position = c(0.75, 0.25),
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
    text = element_text(size = 12),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold")
  ) +
  stat_cor(aes(label = paste(..rr.label..)))

```

# Per Genre
Now we produce the same plot as the overall plot above but for each genre in the dataset.

```{r per_genre, out.width = "100%"}
# The same plot but now for action movies
action = ggplot(aes(x = startYear, y = mean_runtime, colour = n), data = movies[grepl("action", tolower(genres)), .(mean_runtime = mean(runtimeMinutes), n = .N), by = startYear]) +
  geom_line(na.rm = TRUE, size = 1, lineend = "round", linejoin = "mitre") +
  geom_hline(alpha = 0.8, yintercept = 89.75, linetype = 2, size = 1.15, color = "darkgrey") +
  scale_y_continuous(limits = c(0, 280), breaks = c(0, 30, 60, 90, 120), expand = c(0, 5)) +
  scale_x_continuous(expand = c(0, 1)) +
  scale_colour_gradient(
    low = "#E7B800", high = "#FC4E07",
    guide = guide_colorbar(
      title.position = "top", title.hjust = 0.55, nbin = 500, barheight = 0.45, barwidth = 8,
      raster = TRUE
    )
  ) +
  labs(x = "Year", y = "Mean Runtime (in Minutes)", colour = "Number of Movies", title = "Action movies") +
  theme_minimal() +
  theme(
    legend.position = c(0.75, 0.15),
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
    text = element_text(size = 12),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold")
  )

# The same plot but now for romance movies
romance = ggplot(aes(x = startYear, y = mean_runtime, colour = n), data = movies[grepl("romance", tolower(genres)), .(mean_runtime = mean(runtimeMinutes), n = .N), by = startYear]) +
  geom_line(na.rm = TRUE, size = 1, lineend = "round", linejoin = "mitre") +
  geom_hline(alpha = 0.8, yintercept = 89.75, linetype = 2, size = 1.15, color = "darkgrey") +
  scale_y_continuous(limits = c(0, 120), breaks = c(0, 30, 60, 90, 120), expand = c(0, 5)) +
  scale_x_continuous(expand = c(0, 1)) +
  scale_colour_gradient(
    low = "#E7B800", high = "#FC4E07", guide = guide_colorbar(
      title.position = "top", title.hjust = 0.55, nbin = 500,
      barheight = 0.45, barwidth = 8, raster = TRUE
    )
  ) +
  labs(x = "Year", y = "Mean Runtime (in Minutes)", colour = "Number of Movies", title = "Romance movies") +
  theme_minimal() +
  theme(
    legend.position = c(0.75, 0.15),
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
    text = element_text(size = 12),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold")
  )
# The same plot but now for drama movies
drama = ggplot(aes(x = startYear, y = mean_runtime, colour = n), data = movies[grepl("drama", tolower(genres)), .(mean_runtime = mean(runtimeMinutes), n = .N), by = startYear]) +
  geom_line(na.rm = TRUE, size = 1, lineend = "round", linejoin = "mitre") +
  geom_hline(alpha = 0.8, yintercept = 89.75, linetype = 2, size = 1.15, color = "darkgrey") +
  scale_y_continuous(limits = c(0, 120), breaks = c(0, 30, 60, 90, 120), expand = c(0, 5)) +
  scale_x_continuous(expand = c(0, 1)) +
  scale_colour_gradient(
    low = "#E7B800", high = "#FC4E07", guide = guide_colorbar(
      title.position = "top", title.hjust = 0.55, nbin = 500,
      barheight = 0.45, barwidth = 8, raster = TRUE
    )
  ) +
  labs(x = "Year", y = "Mean Runtime (in Minutes)", colour = "Number of Movies", title = "Drama movies") +
  theme_minimal() +
  theme(
    legend.position = c(0.75, 0.15),
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
    text = element_text(size = 12),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold")
  )


plot_grid(action, drama, romance, nrow = 1, ncol = 3, scale = 0.95)

```

# Movies I Watched
Now that we see that although movie runtime has trended upward since 1900, the runtime of movies that were released since I have been born has remained relatively flat. Thankfully I keep extensive records of movies I have watched (although record keeping only started around 2017), so we can check if movies **I have specifically watched** have increased in runtime over time.

```{r personal_movies, out.width = '100%'}
# I have already downloaded my ratings for all movies I have watched in json format
library("jsonlite")
movies_watched = fromJSON("data/movies.json")

setDT(movies_watched)
# convert runtime in minutes to numeric variables
movies_watched$runtime = as.numeric(gsub(r"{\D}", "", movies_watched$runtime))

# convert years to numeric dataset
movies_watched$year = as.numeric(movies_watched$year)

ggplot(aes(x = year, y = mean_runtime, colour = n), data = movies_watched[, .(mean_runtime = mean(runtime), n = .N), by = year]) +
  geom_line(size = 2, lineend = "round", linejoin = "mitre") +
  geom_hline(alpha = 0.8, yintercept = 121.1, linetype = 2, size = 1.15, color = "darkgrey") +
  scale_y_continuous(limits = c(0, 180), breaks = c(0, 30, 60, 90, 120, 150, 180), expand = c(0, 5)) +
  scale_x_continuous(expand = c(0, 1)) +
  scale_colour_gradient(
    low = "#E7B800", high = "#FC4E07", breaks = c(0, 4000, 8000, 12000),
    guide = guide_colorbar(
      title.position = "top", title.hjust = 0.05, nbin = 500, barheight = 0.45, barwidth = 8,
      raster = TRUE
    )
  ) +
  labs(
    x = "Year", y = "Mean Runtime (in Minutes)", colour = "Number of Movies", title = "Mean runtime for all movies I
    have watched", subtitle = "Dotted line is overall mean runtime (121.1 min). Data from Me"
  ) +
  theme_minimal() +
  theme(
    legend.position = c(0.75, 0.25),
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
    text = element_text(size = 12),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold")
  ) +
  stat_cor(aes(label = paste(..rr.label..)))

```

# Conclusion

From all the figures it seems that movie runtime has slightly increased since 1900, and the runtime of movies I have
watched has slightly increased over time but not noticeably.

# Compute Environment

```{r session_info}
sessioninfo::session_info()
```
